Visualisation of Brain Statistics with R-packages \emph{ggseg} and \emph{ggseg3d}

### 2019-12-11 draft_2.tex
revised as discussed

% Pandoc Header
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Visualisation of Brain Statistics with R-package \texttt{ggseg}}

@@

\corrauthor[1]{Athanasia M. Mowinckel}{\href{mailto:a.m.mowinckel@psykologi.uio.no}{\nolinkurl{a.m.mowinckel@psykologi.uio.no}}}
\author[1]{Didac Vidal Piñeiro}

@@

\begin{abstract}
There is an increasing emphasis on visualizing results from statistical analysis in more intuitive ways. Neuroimaging data already have, as a consequence of the source of the data, particular shapes and as such visualization of these data should reflect these same dimensional properties. The ggseg package makes it possible to visualize pre-defined brain segmentations of local/structural or connectivity features in both 2D polygons or 3D mesh. Presenting both relative position and size of segments in addition to statistical summary information aides interpretation and dissemination of results. In this tutorial, we present the data and functions in the ggseg and ggseg3d R-packages for brain atlas visualization. The tutorial highlights the main functions to create brain segmentation plots in R, and also introduces the accompanying package data that makes these visualizations possible.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Visualization is increasingly important for accurate guidance and interpretation of neuroimaging results, as current research is able to generate a high amount of data and outcomes.
For Magnetic Resonance Imaging (MRI), neuroimaging software provides whole-brain information by using many small units of space (\textgreater100.000).
Nonetheless, this data is often grouped and summarized into a limited number of regions using predefined brain parcellation atlases.
Brain parcellations segment the brain into a finite set of meaningful neurobiological components, which reflect one or more brain features either based on local/structural or on connectivity properties (\citet{eickhoff_2018}).
The use of brain atlases is widespread as these facilitate interpretation and minimize the amount of data, hence reducing problems with multiple comparisons.
This facilitates replicability and data sharing in otherwise computationally expensive analyses, which are often performed in specialized software environments such as R (\citet{R}).

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, rather than on other types of graphs (e.g.~bar charts) alone.
The projection of data onto brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and conveys the spatial patterns of the data.
Adopting the grammar of graphics implemented in ggplot2 (\citet{ggplot}), one can plot neuroimaging data directly in R with several tools such as ggBrain (\citet{ggBrain}) and ggneuro (\citet{ggneuro}; see neuroconductor \citeyearpar{neuroconductor} for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The ggseg and ggseg3d -- pluss the complimentary ggsegExtra -- packages include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
Two-dimensional functionality in ggseg is based on polygons and ggplot2-based grammar of graphics (\citet{ggplot}), while the 3D functionality in ggseg3d is based on tri-surface mesh plots and plotly (\citet{plotly}).

Both packages present compiled data sets, tailored functions that allow brain data integration, and plotting, as well as other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; \citet{dkt}), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (\citet{amlien_elaboration_2019}, \citet{WALHOVD20051261}, \citet{Pizzagalli}).
The ggsegExtra package includes a library of precompiled atlases (currently including 13 additional atlases) and it is updated continuously.
See Table \ref{tab:atlasTab} for a summary of the available atlases (as of November 2019).

@@

\caption{\label{tab:atlasTab}Table of currently available atlases in the ggseg, ggseg3d, and the ggsegExtra R-packages. Polygon and mesh refer to 2D and 3D brain atlas representations, respectively}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllll}
\toprule
Title & Item & Mesh & Polygon & citation\\
\midrule
Desikan-Killiany Cortical Atlas & dkt & ggseg3d & ggseg & \citet{dkt}\\
Desterieux cortical parcellations & desterieux & ggsegExtra & None & \citet{desterieux}\\
Freesurfer automatic subcortical segmentation of a brain volume & aseg & ggseg3d & ggseg & \citet{aseg}\\
Genetic topography of brain area morphology & chenAr & None & ggsegExtra & \citet{chen}\\
Genetic topography of brain thickness morphology & chenTh & None & ggsegExtra & \citet{chen}\\
\addlinespace
Harvard-Oxford Cortical atlas & hoCort & None & ggsegExtra & \citet{ho}\\
Parcellation from a midsagittal slice & midsagittal & None & ggsegExtra & \\
Parcellation from JHU & jhu & ggsegExtra & ggsegExtra & \citet{jhu}\\
Parcellation from of white matter & icbm & ggsegExtra & None & \citet{icbm}\\
Parcellation from the Human Connectome Project & glasser & ggsegExtra & ggsegExtra & \citet{glasser}\\
\addlinespace
Schaefer 17 Resting-state Cortical Parcellations & schaefer17 & ggsegExtra & None & \citet{schaefer}\\
Schaefer 7 Resting-state Cortical Parcellations & schaefer7 & ggsegExtra & None & \citet{schaefer}\\
White matter tract parcellations & tracula & ggsegExtra & ggsegExtra & \citet{tracula}\\
Yeo 17 Resting-state Cortical Parcellations & yeo17 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
Yeo 7 Resting-state Cortical Parcellations & yeo7 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
\bottomrule
\end{tabular}}
\end{table}

@@

This tutorial will introduce the \texttt{ggseg}, \texttt{ggseg3d}, and \texttt{ggsegExtra} packages and familiarize the reader with the main functions and the general use of the packages.
The tutorial will focus on the two main functions: \texttt{ggseg()} for plotting 2D polygons and \texttt{ggseg3d()} for plotting 3D brains based on tri-surface mesh plots.

@@

\texttt{ggseg} is the main function for plotting 2D data.
By default, the function automatically plots the DKT atlas (see Figure \ref{fig:init}).
The \texttt{ggseg} function is a wrapper for \texttt{geom\_polygon} from \texttt{ggplot2}, and it can be built upon and combined like any ggplot object.
The image plot consists of a simple brain representation containing no extra information.
Hence, \texttt{ggseg} plots can be easily complemented with any of the available ggplot features and options.
We recommend users to get familiarized with ggplot2 grammar (\citet{ggplot}).
The package is currently only available through github, but we expect to submit to The Comprehensive R Archive Network (\citet{cran}) some time during 2020.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# remotes::install_github("LCBC-UiO/ggseg")}
\KeywordTok{library}\NormalTok{(ggseg)}
\KeywordTok{library}\NormalTok{(tidyverse)}

\CommentTok{# Figure 1}
\KeywordTok{ggseg}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/init-1.pdf}
\caption{\label{fig:init}By default \texttt{ggseg} will plot the dkt atlas in grey shaded polygons.}
\end{figure}

In addition to the standard options for ggplot polygon geoms, the function also has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the `dkt', one can stack the hemispheres, view only the medial or lateral side, choose either one or both hemispheres, or any combination of hemisphere and view.
See Figure \ref{fig:collection}) for examples.
For subcortical atlases, such as the `aseg', the options are more limited but one can often choose between axial, sagittal, and coronal views.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# dkt dark theme}
\NormalTok{p1 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme_dark}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{# dkt classic theme}
\NormalTok{p2 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme_classic}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# dkt medial view}
\NormalTok{med <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# dkt left hemisphere}
\NormalTok{left <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# aseg default theme}
\NormalTok{p3 <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{atlas=}\NormalTok{aseg) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{# dkt left medial alone}
\NormalTok{combo <-}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{,}
      \DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{# Combine plots to Figure 2}
\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot_grid}\NormalTok{(p1, med, combo,  p2, left, p3,}
                   \DataTypeTok{labels =} \KeywordTok{c}\NormalTok{(}\StringTok{"A: dkt - dark"}\NormalTok{, }\StringTok{"B: dkt - medial"}\NormalTok{,}
                              \StringTok{"C: dkt - combo"}\NormalTok{, }\StringTok{"D: dkt - classic"}\NormalTok{,}
                              \StringTok{"E: dkt - left"}\NormalTok{, }\StringTok{"F: aseg"}\NormalTok{),}
                   \DataTypeTok{hjust =} \FloatTok{-.05}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/collection-1.pdf}
\caption{\label{fig:collection}ggseg plots can be used with any ggplot feasture such as standard scales and themes. For cortical atlases, one can supply special ggseg options to determine hemisphere or view, and control how the plots looks like and is organised. \textbf{A:} dkt atlas, stacked with dark theme ; \textbf{B:} dkt with medial view only; \textbf{C:} dkt atlas with only left medial display; \textbf{D:} dkt atlas, stacked, with classic theme; \textbf{E:} dkt atlas with left hemisphere only; \textbf{F:} complete aseg atlas}
\end{figure}

@@

\texttt{ggseg} accepts any argument you can supply to \texttt{geom\_polygon} and therefore is easy to work with for those familiar with ggplot functionality.
Standard arguments like \texttt{fill} that floods the segments with a colour, or \texttt{colour} that colours the edges around the segments are typical arguments to provide to the function either as a single setting value or within the ggplot mapping function \texttt{aes}.
To use color palettes corresponding to those used in the original neuroimaging softwares one can use atlas-specific `brain' palette scales.
See an example in Figure \ref{fig:fill}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Figure 3}
\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ area), }\DataTypeTok{colour=}\StringTok{"black"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_fill_brain}\NormalTok{(}\StringTok{"dkt"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.justification=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{0}\NormalTok{),}
        \DataTypeTok{legend.position=}\StringTok{"bottom"}\NormalTok{,}
        \DataTypeTok{legend.text =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{5}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide_legend}\NormalTok{(}\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/fill-1.pdf}
\caption{\label{fig:fill}Supplying `area' to the fill option in ggseg, will use the column `area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The `dkt' atlas default palette corresponds to the FreeSurferColorLut scheme.}
\end{figure}

Most users will use ggseg to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or brain-cognition relationships across the different brain regions.
Yet, before projecting the statistics onto the segments, we need to explore the structure of the atlas data sets.
The atlas data set structure will help us understand what incoming statistical data needs to look like.
Note that each atlas corresponds to a unique data set.
All data sets have a similar structure and contain key information regarding the atlas, the region names, and the coordinates for the segment polygons.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{head}\NormalTok{(dkt, }\DecValTok{5}\NormalTok{)}
\CommentTok{## # A tibble: 5 x 6}
\CommentTok{##   atlas area               hemi  side    label              ggseg          }
\CommentTok{##   <chr> <chr>              <chr> <chr>   <chr>              <list>         }
\CommentTok{## 1 dkt   superior temporal  left  lateral lh_superiortempor~ <tibble [1,494~}
\CommentTok{## 2 dkt   pre central        left  lateral lh_precentral      <tibble [1,314~}
\CommentTok{## 3 dkt   post central       left  lateral lh_postcentral     <tibble [1,164~}
\CommentTok{## 4 dkt   rostral middle fr~ left  lateral lh_rostralmiddlef~ <tibble [1,194~}
\CommentTok{## 5 dkt   insula             left  lateral lh_insula          <tibble [870 x~}
\end{Highlighting}
\end{Shaded}

>>>>

% Pandoc Header
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Visualisation of Brain Statistics with R-packages \emph{ggseg} and \emph{ggseg3d}}

@@

\corrauthor[1]{Athanasia M. Mowinckel}{\href{mailto:a.m.mowinckel@psykologi.uio.no}{\nolinkurl{a.m.mowinckel@psykologi.uio.no}}}
\author[1]{Didac Vidal-Piñeiro}

@@

\begin{abstract}
There is increased emphasis on visualizing results from statistical analysis in more intuitive ways. Results from imaging data inherently have a spatial component. Commonly used tools to disseminate results (i.e.~bar-charts or similar) ignore this spatial component. A tool that easily integrates the spatial aspect of the data could greatly improve dissemination of results, and also increase efficiency of workflows. The ggseg-packages make it possible to visualize pre-defined brain segmentations as both 2D polygons or 3D meshes. Presenting both the relative position and size of segments in addition to the statistical summary information eases interpretation and dissemination of results. Here we present the main data and functions in the \emph{ggseg} and \emph{ggseg3d} R-packages for brain atlas visualization. The main highlighted functions create brain segmentation plots in R, and datasets included in the packages, as well as in the \emph{ggsegExtra} companion package, fascilitate these visualizations.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

Visualization is increasingly important for accurate guidance and interpretation of neuroimaging results, as current research is able to generate a high amount of data and outcomes.
For Magnetic Resonance Imaging (MRI), neuroimaging software provides whole-brain information by using many small units of space (\textgreater100.000).
Nonetheless, this data is often grouped and summarized into a limited number of regions using predefined brain parcellation atlases.
Brain parcellations segment the brain into a finite set of meaningful neurobiological components, which reflect one or more brain features either based on structural or connectivity properties (\citet{eickhoff_2018}).
The use of brain atlases is widespread as these facilitate interpretation and minimize the amount of data, hence reducing problems with multiple comparisons.
This enables replicability and data sharing in otherwise computationally expensive analyses, which are often performed in specialized software environments such as R (\citet{R}).

MRI data provides good spatial resolution and thus an optimal representation has to respect spatial relationships across regions.
Results from brain atlas analyses are most meaningfully visualized when projected onto a representation of the brain, thus it is desirable that any visual representation respects this relation.
The projection of data onto brain representations provides clear points of reference - especially when the reader is unfamiliar with the atlas - eases readability, guides interpretation, and conveys the spatial patterns of the data.
Adopting the grammar of graphics implemented in \emph{ggplot2} (\citet{ggplot}), one can plot neuroimaging data directly in R with several tools such as ggBrain (\citet{ggBrain}) and ggneuro (\citet{ggneuro}; see neuroconductor \citeyearpar{neuroconductor} for curated neuroimaging packages for R).
Yet these tools display whole-brain image files and are not well-suited for representing brain atlas data.

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The \emph{ggseg} and \emph{ggseg3d} -- pluss the complimentary \emph{ggsegExtra} -- packages include pre-compiled datasets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in \emph{ggseg} is based on polygons and \emph{ggplot2}-based grammar of graphics (\citet{ggplot}), while the 3D functionality in \emph{ggseg3d} is based on tri-surface mesh plots and \emph{plotly} (\citet{plotly}).

Both packages present compiled datasets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DKT; \citet{dkt}), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (\citet{fischl_99}, \citet{dale_99}, \citet{Fischl2000}), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (\citet{amlien_elaboration_2019}, \citet{WALHOVD20051261}, \citet{Pizzagalli}).
The \emph{ggsegExtra} package contains a collection of precompiled atlases (currently 15 additional atlases) and it is updated frequently.
A summary of all available atlases compatible with the \emph{ggseg}-packages as of December 2019 can be found in \ref{tab:atlasTab}.

@@

\caption{\label{tab:atlasTab}Table of currently available atlases in the ggseg, ggseg3d, and the ggsegExtra R-packages. Polygon and mesh refer to 2D and 3D brain atlas representations, respectively}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllll}
\toprule
Title & Item & Mesh & Polygon & citation\\
\midrule
Automated probabilistic reconstruction of white-matter pathways & tracula & None & ggsegExtra & \citet{tracula}\\
Desikan-Killiany Cortical Atlas & dkt & ggseg3d & ggseg & \citet{dkt}\\
Desterieux cortical parcellations & desterieux & ggsegExtra & None & \citet{desterieux}\\
Freesurfer automatic subcortical segmentation of a brain volume & aseg & ggseg3d & ggseg & \citet{aseg}\\
Genetic topography of brain area morphology & chenAr & None & ggsegExtra & \citet{chen}\\
\addlinespace
Genetic topography of brain thickness morphology & chenTh & None & ggsegExtra & \citet{chen}\\
Harvard-Oxford Cortical atlas & hoCort & None & ggsegExtra & \citet{ho}\\
HPC - Multi-modal parcellation of human cerebral cortex & glasser & None & ggsegExtra & \citet{glasser}\\
ICBM white matter parcellation & icbm & ggsegExtra & None & \citet{icbm}\\
JHU parcellation & jhu & None & ggsegExtra & \citet{jhu}\\
\addlinespace
Local-Global 17 Parcellation of the Human Cerebral Cortex & schaefer17 & ggsegExtra & None & \citet{schaefer}\\
Local-Global 7 Parcellation of the Human Cerebral Cortex & schaefer7 & ggsegExtra & None & \citet{schaefer}\\
Parcellation from JHU & jhu & ggsegExtra & None & \citet{jhu}\\
Parcellation from the Human Connectome Project & glasser & ggsegExtra & None & \citet{glasser}\\
White matter tract parcellations & tracula & ggsegExtra & None & \citet{tracula}\\
\addlinespace
Yeo 17 Resting-state Cortical Parcellations & yeo17 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
Yeo 7 Resting-state Cortical Parcellations & yeo7 & ggsegExtra & ggsegExtra & \citet{yeo2011}\\
\bottomrule
\end{tabular}}
\end{table}

@@

This tutorial will introduce the \emph{ggseg}, \emph{ggseg3d}, and \emph{ggsegExtra} packages and familiarize the reader with the main functions and the general use of the packages.
The tutorial will focus on the two main functions: \texttt{ggseg()} for plotting 2D polygons and \texttt{ggseg3d()} for plotting 3D brains based on tri-surface mesh plots.

@@

\emph{ggseg} is the main function for plotting 2D data.
By default, the function automatically plots the DKT atlas (see Figure \ref{fig:init}).
The \texttt{ggseg()}-function is a wrapper for \texttt{geom\_polygon()} from \emph{ggplot2}, and it can be built upon and combined like any \emph{ggplot}-object.
The image plot consists of a simple brain representation containing no extra information.
Hence, \emph{ggseg} plots can be easily complemented with any of the available \emph{ggplot2} features and options.
We recommend users to get familiarized with \emph{ggplot2} (\citet{ggplot}).
The package is currently only available through github, but we expect to submit to The Comprehensive R Archive Network (\citet{cran}) some time during 2020.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# remotes::install\_github("LCBC{-}UiO/ggseg")}
\KeywordTok{library}\NormalTok{(ggseg)}
\KeywordTok{library}\NormalTok{(dplyr)}
\KeywordTok{library}\NormalTok{(tidyr)}

\CommentTok{\# Figure 1}
\KeywordTok{ggseg}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/init-1.pdf}
\caption{\label{fig:init}By default \texttt{ggseg()} will plot the dkt atlas in grey shaded polygons.}
\end{figure}

In addition to the standard options for \emph{ggplot2} polygon geoms, the function also has several options for plotting the main brain representations.
These options are atlas-specific.
For cortical atlases, such as the \texttt{dkt}, one can stack the hemispheres, view only the medial or lateral side, choose either one or both hemispheres, or any combination of hemisphere and view (see Figure \ref{fig:collection} for examples).
For subcortical atlases, such as the \texttt{aseg} atlas, the options are more limited but one can often choose between axial, sagittal, and coronal views.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# dkt dark theme}
\NormalTok{p1 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme\_dark}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt classic theme}
\NormalTok{p2 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{position =} \StringTok{"stacked"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme\_classic}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt medial view}
\NormalTok{med <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt left hemisphere}
\NormalTok{left <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# aseg default theme}
\NormalTok{p3 <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{atlas=}\NormalTok{aseg) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title =} \StringTok{" "}\NormalTok{)}

\CommentTok{\# dkt left medial alone}
\NormalTok{combo <{-}}\StringTok{ }\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{view =} \StringTok{"medial"}\NormalTok{,}
               \DataTypeTok{hemisphere =} \StringTok{"left"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{" "}\NormalTok{)}

\CommentTok{\# Combine plots to Figure 2}
\NormalTok{cowplot}\OperatorTok{::}\KeywordTok{plot\_grid}\NormalTok{(p1, med, combo,  p2, left, p3,}
                   \DataTypeTok{labels =} \KeywordTok{c}\NormalTok{(}\StringTok{"A: dkt {-} dark"}\NormalTok{, }\StringTok{"B: dkt {-} medial"}\NormalTok{,}
                              \StringTok{"C: dkt {-} combo"}\NormalTok{, }\StringTok{"D: dkt {-} classic"}\NormalTok{,}
                              \StringTok{"E: dkt {-} left"}\NormalTok{, }\StringTok{"F: aseg"}\NormalTok{),}
                   \DataTypeTok{hjust =} \FloatTok{{-}.05}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/collection-1.pdf}
\caption{\label{fig:collection}\emph{ggseg} plots can be used with any \emph{ggplot2} feature such as standard scales and themes. For cortical atlases, one can supply special \emph{ggseg} options to determine hemisphere or view, and control how the plots looks like and is organised. \textbf{A:} \texttt{dkt} atlas, stacked with dark theme ; \textbf{B:} \texttt{dkt} with medial view only; \textbf{C:} \texttt{dkt} atlas with only left medial display; \textbf{D:} \texttt{dkt} atlas, stacked, with classic theme; \textbf{E:} \texttt{dkt} atlas with left hemisphere only; \textbf{F:} complete \texttt{aseg} atlas}
\end{figure}

@@

\texttt{ggseg()} accepts any argument you can supply to \texttt{geom\_polygon()}, and therefore is easy to work with for those familiar with \emph{ggplot2} functionality.
Standard arguments like \texttt{fill} that floods the segments with a colour, or \texttt{colour} that colours the edges around the segments are typical arguments to provide to the function either as a single setting value or within the \emph{ggplot2} mapping function \texttt{aes}.
To use color palettes corresponding to those used in the original neuroimaging softwares one can use atlas-specific `brain' palette scales (Figure \ref{fig:fill}).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Figure 3}
\KeywordTok{ggseg}\NormalTok{(}\DataTypeTok{mapping=}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ area), }\DataTypeTok{colour=}\StringTok{"black"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale\_fill\_brain}\NormalTok{(}\StringTok{"dkt"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.justification=}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{0}\NormalTok{),}
        \DataTypeTok{legend.position=}\StringTok{"bottom"}\NormalTok{,}
        \DataTypeTok{legend.text =} \KeywordTok{element\_text}\NormalTok{(}\DataTypeTok{size =} \DecValTok{5}\NormalTok{)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{guides}\NormalTok{(}\DataTypeTok{fill =} \KeywordTok{guide\_legend}\NormalTok{(}\DataTypeTok{ncol =} \DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\centering
\includegraphics{draft_2_files/figure-latex/fill-1.pdf}
\caption{\label{fig:fill}Supplying `area' to the fill option in \texttt{ggseg()}, will use the column `area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The \texttt{dkt} atlas default palette corresponds to the FreeSurferColorLut scheme.}
\end{figure}

Most users will use \texttt{ggseg()} to display - using a color scale - some descriptive or inferential statistics, such as mean thickness or brain-cognition relationships across the different brain regions.
Yet, before projecting the statistics onto the segments, one should explore the structure of the atlas datasets.
The atlas dataset structure will help users understand what incoming statistical data needs to look like.
Note that each atlas corresponds to a unique dataset.
All datasets have a similar structure and contain key information regarding the atlas, the region names, and the coordinates for the segment polygons.

In any atlas, the column `label' is particularly useful for combining the data of interest with the \emph{ggseg}-polygons.
The column `label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in \emph{ggseg} is in a long format - that is each region has its own row - and any data of interest needs to be in this same format.
Often datasets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged in order to work with ggseg.
See below an example of wide-to-long conversion.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Create a mock freesurfer output}
\NormalTok{freesurfer\_stats <{-}}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}
  \DataTypeTok{id =} \KeywordTok{c}\NormalTok{(}\DecValTok{10}\OperatorTok{:}\DecValTok{12}\NormalTok{),}
  \DataTypeTok{lh\_superiortemporal =} \KeywordTok{c}\NormalTok{(}\FloatTok{3.32}\NormalTok{, }\FloatTok{4.1}\NormalTok{, }\FloatTok{3.5}\NormalTok{),}
  \DataTypeTok{lh\_precentral =} \KeywordTok{c}\NormalTok{(}\FloatTok{2.3}\NormalTok{, }\FloatTok{2.5}\NormalTok{, }\FloatTok{2.1}\NormalTok{),}
  \DataTypeTok{lh\_rostralmiddlefrontal =} \KeywordTok{c}\NormalTok{(}\FloatTok{3.3}\NormalTok{, }\FloatTok{3.2}\NormalTok{, }\FloatTok{3.1}\NormalTok{)}
\NormalTok{)}
\NormalTok{freesurfer\_stats}
